---
title: "PCA Introduction"
author: ""
date: "2022-10-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, message = FALSE,warning = FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
library(readxl)
library(tidyverse)
library(janitor)
library(rgl)
library(car)
library(tidyverse)
library(ggfortify)
library(ggforce)
knitr::knit_hooks$set(webgl = hook_webgl)
cbPalette <- c("#E69F00", "#009E73", "#0072B2", "#CC79A7") # custom color-blind friendly palette
```

You'll need to run the following code (in the console, *not* in RMarkdown):

```
install.packages("rgl")
install.packages("car")
install.packages("ggfortify")
install.packages("ggforce")
install.packages("MASS")
```

Then run the R script `simulate_data.R` to simulate data for this lab exercise.

```{r}
source("simulate_data.R")
```

This script created a variable named `X` with 3 plant trait columns: Nitrogen (`Nitrogen`), leaf mass per area (`LMA`), and Water Content (`Water`):

```{r}
head(X)
```

The data are also stored in a data frame called `dat`, but `dat` also has a column named `Type`, which cateogrizes plants as "Dessert Annual", "Dessert Perennial", "Temperate Annual", or "Temperate Perennial".

```{r}
head(dat)
```

Let's see how the variables are correlated with one another:

```{r}
cor(X) # r
cor(X)^2 # r-squared
```

We see that Nitrogen and LMA are strongly negatively correlated:

```{r}
dat %>% 
  ggplot(mapping=aes(x = Nitrogen,y = LMA,color = Type)) + 
  geom_point(color = cols,size = 3) + 
  theme_classic()
```

Next we visualize Nitrogen vs Water, which has a very weak correlation.

```{r}
dat %>% 
  ggplot(mapping=aes(x = Nitrogen,y = Water,color = Type)) +
  geom_point(color = cols,size = 3) + 
  theme_classic()
```

Finally, visualize LMA vs Water Content, which is moderately positively correlated.

```{r}
dat %>% 
  ggplot(mapping=aes(x = LMA,y = Water,color = Type)) + 
  geom_point(color = cols,size = 3) + 
  theme_classic()
```

We can view the 3 traits simultaneously in a 3D plot:

```{r webgl=TRUE}
scatter3d(x = X[,"Nitrogen"],y = X[,"LMA"],z = X[,"Water"],
          point.col = cols,
          surface.alpha=.00,
          residuals = FALSE,
          sphere.size=1.5,
          fogtype="none",
          xlab = "Nitrogen",
          ylab = "LMA",
          zlab = "Water")
```

## Principal Component Analysis (PCA)

Because these traits are correlated, there is redundant information: if you know Nitrogen, you have a pretty good idea of LMA, and if you know LMA, you have a decent idea of Water Content.

In other words, we should be able to summarize an individual plant's Nitrogen content, LMA, and Water Content with just two variables. Principal component analysis (PCA) is a common way to summarize the traits of several variables with a small number of summary variables.

The following code runs PCA on the data `X`, and then prints the summarized results.

We see the first principal component (PC1) explains nearly three-quarters of the variation in the data! PC2 explains another quarter, and PC3 doesn't tell us much of anything.

```{r}
pca <- prcomp(X,center = TRUE,scale. = TRUE)
summary(pca)
```

We can also see how correlated (r and R^2) the traits are with each PC:

```{r}
cor(X,pca$x) # r
cor(X,pca$x)^2 # r-squared
```

Visualize PC1 and PC2 (together they convey almost all of the information about the 3 traits):

```{r}
autoplot(object = pca,
         data = dat,
         x = 1,
         y = 2,
         loadings = TRUE,
         loadings.label = TRUE) + 
  geom_point(color = cols,size = 3) + 
  theme_classic()
```

Visualizing PC1 and PC3 doesn't provide any new information.

```{r}
autoplot(object = pca,
         data = dat,
         x = 1,
         y = 3,
         loadings = TRUE,
         loadings.label = TRUE) + 
  geom_point(color = cols,size = 3) + 
  theme_classic()
```

Nor does visualizing PC2 and PC3.

```{r}
autoplot(object = pca,
         data = dat,
         x = 2,
         y = 3,
         loadings = TRUE,
         loadings.label = TRUE) + 
  geom_point(color = cols,size = 3) + 
  theme_classic()
```

## Visualing by groups

Now let's look at the data in context of their habitat. First, Nitrogen vs Water Content:

```{r}
dat %>% ggplot(mapping=aes(x = Nitrogen,y = Water,color = Type)) +   geom_mark_ellipse(mapping = aes(fill = Type),expand = unit(0.5,'mm')) + 
  geom_point(mapping = aes(color = Type),size = 3) +
  scale_fill_manual(values = cbPalette) + 
  scale_color_manual(values = cbPalette) + 
  theme_classic()
```

Nitrogen vs LMA:

```{r}
dat %>% ggplot(mapping=aes(x = Nitrogen,y = LMA,color = Type)) +   geom_mark_ellipse(mapping = aes(fill = Type),expand = unit(0.5,'mm')) + 
  geom_point(mapping = aes(color = Type),size = 3) +
  scale_fill_manual(values = cbPalette) + 
  scale_color_manual(values = cbPalette) + 
  theme_classic()
```

LMA vs Water:

```{r}
dat %>% ggplot(mapping=aes(x = LMA,y = Water,color = Type)) +   geom_mark_ellipse(mapping = aes(fill = Type),expand = unit(0.5,'mm')) + 
  geom_point(mapping = aes(color = Type),size = 3) +
  scale_fill_manual(values = cbPalette) + 
  scale_color_manual(values = cbPalette) + 
  theme_classic()
```

3D plot:

```{r webgl=TRUE}
scatter3d(x = X[,"Nitrogen"],y = X[,"LMA"],z = X[,"Water"],
          point.col = cbPalette[dat$Type],
          surface.alpha=.00,
          residuals = FALSE,
          sphere.size=1.5,
          fogtype="none",
          xlab = "Nitrogen",
          ylab = "LMA",
          zlab = "Water")
```

PCA with groups:

```{r}
autoplot(pca,data = dat,loadings = TRUE,loadings.label = TRUE) +
  geom_mark_ellipse(mapping = aes(fill = Type),expand = unit(0.5,'mm')) + 
  geom_point(mapping = aes(color = Type),size = 3) +
  scale_fill_manual(values = cbPalette) + 
  scale_color_manual(values = cbPalette) + 
  theme_classic()
```

## Sunflower HPLC Data

```{r}
library(readxl)
library(tidyverse)
library(janitor)

data <- read_excel("data/Mason_S1.xlsx", 
    sheet = "Population_LSMeans", skip = 2) %>% janitor::clean_names()
data$growth_form <- as.factor(data$growth_form)
data <- data %>% filter(complete.cases(data %>% select(starts_with("Compound"))))
data <- data %>% mutate(across(starts_with("Compound"),function(X) X))

compounds <- data %>% 
  select(species,growth_form,number_compounds,starts_with("Compound"))
compounds_pca <- prcomp(x = compounds %>% select(where(is_double)),center=TRUE,scale=TRUE)

# which PCs are potentially useful?
round(summary(compounds_pca)$importance,2)

# what compounds are most correlated with the most informative PCs?
round(cor(compounds %>% select(where(is_double)),compounds_pca$x)^2,2)[,1:4]
```

```{r}
scatter3d(x = data$compound_76,y = data$compound_369,z = data$number_compounds,
          point.col = cbPalette[compounds$growth_form],
          surface.alpha=.00,
          residuals = FALSE,
          sphere.size=1.5,
          fogtype="none",
          xlab = "Compound 76",
          ylab = "Compound 369",
          zlab = "Number Compounds")
```

```{r}
autoplot(compounds_pca,data = compounds,loadings = TRUE,loadings.label = TRUE) +
  geom_mark_ellipse(mapping = aes(fill = growth_form),alpha=.3,expand = unit(0.5,'mm')) + 
  geom_point(mapping = aes(color = growth_form),size = 3) +
  scale_fill_manual(values = cbPalette) + 
  scale_color_manual(values = cbPalette) + 
  theme_classic()
```